# Feb 17: **ASSIGNMENT** Stage-structured population dynamics

Class will be in CSF 2218

ASSIGNMENT 3 is due on March 3.

Your assignment will be to write an R markdown file that does these same calculations but for a different projection matrix, $A$ (further details are provided after these sample calculations).

Install and load the `popbio` package.

```{r}
require("popbio")
```

Define our projection matrix (this is the one from class on Feb 16, but with 0.2 in row 2, column 2, so this is a stage-structured, rather than and age-structured, projection matrix, where there is a 0.2 probability of surviving to the next year when older than 1 year. The stages are 0-1 years old and $>1$ year old):
```{r}
A = matrix(c(0,2,0.5,0.2), 2,2, byrow=TRUE)
A
```

Type `A` into the console to check that A has the correct number of rows and columns and the numerical entries in the correct place.

Read about the function `pop.projection()`. (One way to do this is to type `?pop.projection` into the Console. The details will appear in the _Help_ pane).

Let us first try to do one iteration (i.e., one future time step) for the projection matrix, when we have the initial values $n = c(10,5)$. This means that initially, i.e., at $t=0$, we have 10 individuals aged 0-1 years, and 5 individuals aged more than 1 year. Because this is in matrix notation, we need to write the initial number of individuals as a vector, i.e. using the $c()$ function.

Let's do this:

```{r}
n0 = c(10,5)
n1 = pop.projection(A, n0, 2)
n1$stage.vectors
```

Here the print out is the number of individuals in each of the stages at time 0 and time 1. You should be able to verify, by doing the matrix multiplication, that $[10, 6]$ is the correct stage-structured population size for $t=1$, for the projection matrix, $A$, when the intial population size is $[10,5]$.

Now, let us continue to estimate the stage-structured population size further into the future:

```{r}
n20 = pop.projection(A, n0, 21)
n20$stage.vectors
```
That output is hard to read and understand so let's graph the result. To graph in `ggplot()` we will need to reformat this output as a dataframe:

```{r}
data = data.frame(time = as.numeric(colnames(n20$stage.vectors)), pop1 = n20$stage.vectors[1,], pop2 = n20$stage.vectors[2,])
data
```

Running the `data` command in the console checks that `data` looks correct when reformatted as a dataframe.

```{r}
require(ggplot2)
g1 = ggplot(data, aes(x=time))+
  geom_line(aes(y=pop1), col = "red")+
  geom_line(aes(y=pop2), col = "blue")+
  ylab("number of ind. in each stage")
g1
```

It is interesting to consider the per capita change in the population size (within a stage) for each time step. We calculate this as:

\[ \lambda_t = 1+\frac{N_{i,t+1} - N_{i,t}}{N_{i,t}} \qquad \mbox{for $t= \{0, 1, \dots, T-1\}$.} \]

Let's do this for each of the stages and plot the result:

```{r}
n1t1 = tail(data$pop1,-1)
n1t = head(data$pop1,-1)
lambda1 = 1+(n1t1 - n1t)/n1t

n2t1 = tail(data$pop2,-1)
n2t = head(data$pop2,-1)
lambda2 = 1+(n2t1 - n2t)/n2t

data2 = data.frame(time = head(data$time,-1), lambda1, lambda2)
g2 = ggplot(data=data2, aes(x=time))+
  geom_line(aes(y=lambda1), col = "red")+
  geom_line(aes(y=lambda2), col = "blue")+
  ylab("lambda_t")+
  ggtitle("per capita change in population size for each stage")
g2
```

This plot has two interesting features:

- eventually the per capita change in population size becomes constant, and

- the constant value is the same for both stages.

In fact, both stages are growing exponentially, and in the limit as $t\to \infty$ the exponential growth rate for both stages is the value that this plot is converging to (i.e., a little more than 1.1).

This has been calculated for us by the `pop.projection()` function (the final per capita change for 20 iterations). Let's recover the value:

```{r}
n20$lambda
```
Super! As we expected: a little more than 1.1.

In fact, the exponential growth rate, for both stages, in the limit as $t \to \infty$ is the _dominant eigenvalue_ of $A$. Let's find the eigenvalues of $A$:

```{r}
eigen(A)$values
```

Because $A$ is a 2 row by 2 column matrix, the _dominant eigenvalue_ is the eigenvalue that is largest in absolute value. Since $|1.105| > |-0.905| = 0.905$ the dominant eigenvalue is $1.105$.

Would we get a closer $\lambda$ value if we did more iterations (i.e. more time steps). Let's try:

```{r}
n100 = pop.projection(A,n0,101)
n100$lambda
eigen(A)$values[1]
```

Let's consider another interesting feature of our stage-structured population dynamics. What is the proportion of age 0-1 year olds and the proportion of $>1$ year olds in the population?

Let's make graph:

```{r}
g3 = ggplot(data, aes(x=time))+
  geom_line(aes(y = pop1/(pop1+pop2)), col = "red")+
  geom_line(aes(y = pop2/(pop1+pop2)), col = "blue")+
  ylim(c(0.3, 0.7))+
  ylab("stage proportions")
g3
```

Super interesting! Although both stages grow exponentially, the _proportion_ of the population eventually reaches a constant value with a little more than 35% of the population over 1 year old.

We can recover this from the `pop.projection()` function too:

```{r}
n20$stable.stage
```
This says that after 20 iterations (time steps) we had 64.4% of the population 0-1 years old and 35.6% of the population >1 year old.

In terms of the projection matrix, $A$, this stable stage structure (the proportion of the stages as $t \to \infty$) is the _right eigenvector_ associated with the dominant eigenvalue.

Let's find this:

```{r}
eigen(A)
```

This output states that the _right eigenvector_ associated with the dominant eigenvalue (recall that the dominant eigenvalue is 1.105) is $[-0.875, -0.484]$. To complete our calculation, we need to normalize the right eigenvector so that it's elements sum to 1. We do this by dividing by its sum:

```{r}
eig.vec = eigen(A)$vectors[,1]
norm.eig.vec = eig.vec/sum(eig.vec)
norm.eig.vec
```

Yay! We have the result we were expecting: over time, eventually, the percentage of the population that are aged 0-1 years becomes closer and closer to 64.4%. Therefore, the right eigenvector associated with the dominant eigenvalue tells as the stable stage structure (i.e. the proportion of individuals in each stage as $t \to \infty$.

\subection{Assignment 3}
You assignment is to write an R markdown file where you do all these same calculations and graphs but for a different projection matrix, $A$.

In addition, your projection matrix must either be:

- a 3-row by 3-column Leslie matrix; or

- a 2-row by 2-column Leslie matrix where the population
_decreases_ exponentially (since this example was exponential increase).

Here is an example of writing a 3 $\times$ 3 matrix:

```{r}
A = matrix(c(1,2,3,4,5,6,7,8,9), 3,3, byrow=TRUE)
```

Note that for Leslie matrix only the elements in the first row can be larger than 1, and all other elements are stage transition probabilities that need to be $\geq 0$ and $\leq 1$.

To achieve a matrix $A$ where the population decreases exponentially you need to choose low reproduction rates and survivorship probabilities.

Provide sensible axes names and titles for your graphs. In your write-up comment when values shown in graphs, and as calculated using functions have the same values as expected. It does not need to be a long write-up, but I need to see that you understand that the asymptotic ($t\to \infty$) properties of the graphs can be calculated as the eigenvalues and eigenvectors of the projection matrices.